\chapter{Background}
\label{ch:background}

\section{Classical Machine Learning}

Machine learning (ML) is a field of artificial intelligence devoted to building algorithms that can learn from data patterns, enabling systems to predict and make decisions with minimal human intervention.
ML algorithms are typically divided into four broad categories: supervised learning, unsupervised learning, semi-supervised learning, and reinforcement learning~\cite{shalev-shwartz2014understanding}.

\begin{itemize}
    \item \textbf{Supervised learning}: involves training models on labeled datasets by learning a function that maps input features to output labels.
    \item \textbf{Unsupervised learning}: finds patterns and structures in unlabeled data, as seen in clustering or dimensionality reduction algorithms.
    \item \textbf{Semi-supervised learning}: combines small amounts of labeled data with larger unlabeled datasets, enhancing efficiency in domains with limited annotation.
    \item \textbf{Reinforcement learning}: guides an agent to optimal behavior through trial and error, rewarding desirable actions.
\end{itemize}

A few representative machine learning models are~\cite{sarker2021machine}:
\begin{itemize}

    \item \textbf{Linear Regression}: Used for predicting continuous variables and modeling the relationship between features and output.
        It retains mathematical simplicity while remaining a crucial benchmark in both research and forecasting.

    \item \textbf{Decision Trees}: Represent decisions with branching structures, offering interpretability and handling mixed-type features.

    \item \textbf{Random Forests}: Ensembles of decision trees, reducing variance and improving generalization.

    \item \textbf{Support Vector Machines (SVM)}: Classifies data by finding the optimal separating hyperplane, being effective in high-dimensional settings.

    \item \textbf{K-Nearest Neighbors (KNN)}: Classifies points based on the “nearest” labeled examples.

    \item \textbf{Gradient Boosting Frameworks (XGBoost, LightGBM, CatBoost)}: Ensemble methods that extend predictive power, speed, and robustness by aggregating the outputs of numerous decision trees to make predictions.
        XGBoost is renowned for its scalability and performance in structured/tabular data~\cite{chen2016xgboost}.
        LightGBM excels at speed and memory efficiency~\cite{ke2017lightgbm}.
        CatBoost handles categorical variables and is competitive in ranking and classification tasks~\cite{prokhorenkova2018catboost}.
\end{itemize}

\section{Deep Learning}
Deep learning is a subset of machine learning that uses artificial neural networks to model and solve complex problems, inspired by the structure and function of biological neurons~\cite{lecun2015deep}.
These networks are composed of multiple layers of neurons, each layer processing the output of the previous layer to extract increasingly complex features.
A typical neural network comprises an input layer, one or more hidden layers, and an output layer.

A few representative deep learning models are:
\begin{itemize}
    \item \textbf{Feedforward Neural Networks (Multi-Layer Perceptrons)}: The core architecture for classifying or regressing fixed-size inputs.
    \item \textbf{Convolutional Neural Networks (CNNs)}: Specialized for spatial relationships, such as image data.
    \item \textbf{Recurrent Neural Networks (RNNs)}: Designed for sequential data, such as text or time series.
    \item \textbf{Transformer Models}: Modern architectures excelling in tasks ranging from natural language understanding to vision.

\end{itemize}

Neural networks are flexible and scalable but can be data-hungry, prone to overfitting, and require careful regularization and tuning to achieve optimal results.

\section{Model Performance Metrics}

Model evaluation is essential for assessing predictive reliability.
Key metrics for regression models include~\cite{bishop2006pattern}:

\begin{itemize}
    \item \textbf{Mean Absolute Error (MAE)}:
        \[
            MAE = \tfrac{1}{n}\sum_{i=1}^n \lvert y_i - \hat{y}_i \rvert \, .
        \]

        This metric conveys the average absolute prediction error, often valued for its interpretability and resistance to outliers.

    \item \textbf{Mean Absolute Percentage Error (MAPE)}:
        \[
            MAPE = \tfrac{100}{n}\sum_{i=1}^n \frac{|y_i - \hat{y}_i|}{|y_i|} \, .
        \]

        MAPE expresses error as a percentage relative to true values, facilitating comparisons across diverse scales. However, MAPE can be unstable when true values approach zero.
        Other metrics (RMSE, $R^2$) complement these in providing insight into the variance, fit, and reliability of regression and classification models.

    \item \textbf{Root Mean Squared Error (RMSE)}:
        \[
            RMSE = \sqrt{\tfrac{1}{n}\sum_{i=1}^n (y_i - \hat{y}_i)^2} \, .
        \]

        RMSE measures the average magnitude of the prediction errors (residuals) between predicted and actual values in regression tasks.
        It is more sensitive to large errors than MAE due to squaring, penalizing outliers heavily.

    \item $\mathbf{R^2}$ (coefficient of determination):
        \[
            R^2 = 1 - \frac{\sum_{i=1}^n (y_i - \hat{y}_i)^2}{\sum_{i=1}^n (y_i - \bar{y})^2} \, .
        \]

        $R^2$ measures the proportion of variance in the target variable explained by the model, ranging from 0 to 1.
        Values near 1 indicate strong fit, while $R^2 \!\leq\! 0$ means the model performs worse than a constant baseline.

\end{itemize}

\section{Regularization, Overfitting, and Underfitting}
Models risk overfitting—memorizing the training set at the cost of poor generalization—when they become excessively complex.
Underfitting happens when the model is too simple, missing essential patterns.

Regularization combats overfitting by penalizing complexity:

\begin{itemize}

    \item \textbf{L1 Regularization (Lasso)}: Encourages sparsity, zeroing non-critical coefficients.
    \item \textbf{L2 Regularization (Ridge)}: Shrinks coefficients smoothly, preferred for stability.

\end{itemize}

In neural networks, additional techniques such as dropout, early stopping, and weight decay further mitigate overfitting.

\section{Cross-Validation}

Cross-validation is the gold standard for robust model evaluation.
By partitioning data into training and validation folds, cross-validation (often k-fold) averages performance over multiple splits, reducing variance and protecting against accidental overfitting to a single dataset segmentation.
Model selection and hyperparameter tuning typically leverage cross-validation to ensure real-world generalizability

Some common cross-validation techniques are:

\begin{itemize}
    \item \textbf{K-Fold Cross-Validation}: Partition the data into k equal folds, and then train on k-1 folds and validate on the remaining fold.

    \item \textbf{Stratified K-Fold Cross-Validation}: Partition the data into k equal folds, and then train on k-1 folds and validate on the remaining fold.
        The folds are stratified, meaning that the distribution of the target variable is kept the same in each fold.

    \item \textbf{Leave-One-Out Cross-Validation}: Partition the data into n equal folds, and then train on n-1 folds and validate on the remaining fold.

    \item \textbf{Repeated K-Fold Cross-Validation}: Repeat K-Fold Cross-Validation n times, and then average the performance over the n repetitions.
\end{itemize}

\section{Model Selection and Hyperparameter Tuning}

Selecting the right model and its hyperparameters is a critical decision in the ML workflow.

\begin{itemize}
    \item \textbf{Model selection} often involves comparing candidate models (such as linear regression, decision tree, or boosting algorithms) using performance metrics evaluated on a validation set or through cross-validation.
    \item \textbf{Hyperparameter tuning} refers to optimizing parameters that govern model behavior but are not learned during training (e.g., learning rate, tree depth, regularization strength, number of layers in a neural network).
        Grid search systematically explores combinations, while random search samples the setup space.
        Advanced approaches such as Bayesian optimization build probabilistic models to efficiently home in on good hyperparameter settings.
\end{itemize}

\section{Continuous Learning}
Continuous learning, also known as continual or lifelong learning, refers to a paradigm in machine learning where models incrementally update their knowledge by learning from new data streams over time, rather than being trained solely once on a static dataset.
In contrast to traditional retrain-from-scratch approaches, continuous learning enables models to adapt to evolving environments, integrate recent information, and remain relevant as data changes.
Key strategies include incremental learning (updating parameters gradually), transfer learning, and experience replay.

A major challenge for continuous learning is preventing ``catastrophic forgetting'', the loss of knowledge about previously learned tasks when adapting to new data~\cite{parisi2019continual}.
Effective approaches balance learning new patterns (flexibility) with retaining established skills (stability).

\section{MLOps}
MLOps, short for Machine Learning Operations, is a discipline at the intersection of machine learning and DevOps, focused on standardizing, automating, and streamlining the end-to-end lifecycle of machine learning models in production environments.
Its goal is to bridge the gap between data science and IT operations, helping organizations move ML prototypes from research to production efficiently and reliably.

Core MLOps principles include~\cite{tabassam2023mlops}:
\begin{itemize}
    \item \textbf{Versioning}: Tracking code, data, models, and experiment metadata to ensure reproducibility and traceability across iterations and deployments.
    \item \textbf{Automation}: Implementing CI/CD (continuous integration/continuous delivery) pipelines for data ingestion, model training, validation, deployment, and rollback, reducing manual workload and errors.
    \item \textbf{Monitoring}: Continuously assessing live models for data drift, concept drift, and performance degradation, enabling timely retraining, adaptation, and governance.
    \item \textbf{Collaboration}: Facilitating teamwork between data scientists, engineers, and domain experts with standardized workflows, shared environments, and modular architectures.
    \item \textbf{Governance and Security}: Enforcing access controls, audit trails, and compliance with regulatory standards, especially in sensitive domains.
\end{itemize}

MLOps empowers organizations to deploy ML solutions at scale, react to switching data environments, and maintain high service quality and reliability.
Challenges involve cultural change, tool integration, and evolving infrastructure requirements.
Prominent MLOps platforms include cloud-native solutions (AWS SageMaker, Google Vertex AI), as well as open-source toolkits like MLflow and Kubeflow.

\section{Concept Drift}

Concept drift is the phenomenon where the statistical properties of the target variable change over time, leading to reduced model accuracy~\cite{lu2018learning}.
This effect is especially critical in streaming and non-stationary environments, such as online monitoring, finance, or cybersecurity.

Concept drift manifests in various forms:
\begin{itemize}
    \item \textbf{Sudden (abrupt)}: rapid changes in data distribution;
    \item \textbf{Gradual}: slow transitions between concepts;
    \item \textbf{Recurring (recurrent)}: cyclic or seasonal patterns;
    \item \textbf{Incremental}: stepwise, small alterations over time.
\end{itemize}

Detection strategies are categorized as error-rate monitoring (e.g., DDM, EDDM, HLFR), statistical tests, and ensembles.
Drift detectors can monitor prediction error, data distribution, or model performance metrics.
Notable techniques include Hierarchical Hypothesis Testing (HHT), Linear Four Rates (LFR), and ensemble detectors, each with distinct strengths relevant to the type and speed of drift.

Accurate and timely drift detection is paramount for maintaining predictive accuracy and reliability.
Systems often employ sliding windows or reference distributions to test for shifts, and performance degradation often triggers retraining and adaptation processes.

Concept drift adaptation strategies help machine learning models stay accurate as data changes over time.
The main approaches include frequent retraining (updating the model on new data), online learning (incremental updates with each new sample), window-based methods (using only recent data for training), and ensemble methods (combining several models and favoring those performing best on recent data).
Some strategies rely on drift detectors to trigger adaptation, while others update models continuously regardless of detected drift.
The goal is to respond quickly to changes, maintaining reliable predictions in evolving environments.

\section{Key Trade-offs in Drift Detection and Adaptation}

Key trade-offs in drift detection and adaptation include~\cite{zliobaite2014survey}:

\begin{itemize}
    \item \textbf{Sensitivity vs. False Alarms}: More sensitive detectors can spot small or early changes but risk frequent false positives.
        Less sensitive thresholds may delay detection, potentially degrading predictive performance.

    \item \textbf{Label Availability}: Some detectors require true labels to estimate error rates, limiting their use in unsupervised or delayed-label scenarios.
        Unsupervised methods can work on input distributions alone but often have less precision.

    \item \textbf{Latency}: Maintaining low-latency detection is crucial for real-time systems but may require sacrificing some detection accuracy or threshold tightness to avoid delays.

    \item \textbf{Computational Cost}: More advanced or frequent drift checks, especially those recalculating statistics or performing window comparisons, can significantly raise resource requirements, impacting scalability in high-volume streams.
\end{itemize}

Carefully balancing these trade-offs is essential when choosing drift detectors and adaptation strategies for different operational contexts, such as real-time monitoring, batch analytics, or resource-constrained environments.

\section{SUMO Simulator}

SUMO (Simulation of Urban MObility) is an open-source, microscopic and continuous multi-modal traffic simulation suite designed to model and analyze the movement of individual vehicles, pedestrians, and various transportation modes on large-scale road networks~\cite{sumo2024}.
Each agent, be it a car, bus, bicycle, or pedestrian is simulated explicitly, with unique routes, behaviors, and interactions, allowing for fine-grained analyses of traffic dynamics and management strategies~\cite{espinosa2021traffic}.

Core functionalities and capabilities include:
\begin{itemize}
    \item \textbf{Microscopic Simulation}: Models each vehicle and pedestrian individually, supporting realistic behaviors such as lane changes, right-of-way rules, and traffic light interactions.
    \item \textbf{Multi-modal Traffic}: Simulates not only private cars but also public transport, bicycles, pedestrians, and intermodal trips within the same environment.
    \item \textbf{Flexible Network Import}: Supports importing road networks from OpenStreetMap, commercial formats (VISUM, Vissim, NavTeq), and others, or allows users to create synthetic networks.
    \item \textbf{Demand Generation}: Generates vehicle flows using origin-destination matrices, real traffic counts, or synthetic (random and virtual population-based) demand models.
    \item \textbf{Traffic Management}: Enables the testing and evaluation of traffic lights, routing policies, eco-aware navigation, and urban policies before real-world deployment.
    \item \textbf{Real-time Control and Interoperability}: Offers APIs like TraCI for live, programmatic control of the simulation and integration with external tools, such as vehicle communication (C2X) simulators.
    \item \textbf{Comprehensive Visualization}: Provides tools for route finding, visualization and detailed performance monitoring.
    \item \textbf{Extensive Simulation Outputs}: Generates vehicle trajectories, trip statistics, traffic states, environmental metrics, and more in XML or CSV, enabling detailed analysis.
    \item \textbf{High Performance and Scalability}: Efficiently manages large networks (10,000+ edges; 100,000+ vehicles) and supports parallel simulations on various platforms (Windows, Linux, macOS).
    \item \textbf{Open Source and Extensible}: Freely available and extensible, with a large user and developer community for enhancements, custom models, and research applications.
\end{itemize}

These features make SUMO an established tool for transportation planning, intelligent transportation system development, traffic light optimization, and research on smart mobility solutions worldwide.

\section{Containerization and Orchestration}
\textbf{Docker} is a widely adopted containerization platform that allows the packaging of software and its dependencies into isolated units called containers~\cite{merkel2014docker}.
This isolation ensures that code runs the same way regardless of the underlying environment, solving the notorious ``it works on my machine'' problem.
Docker containers are lightweight, highly portable, and enable versioned environments.
Academic use cases include reproducible computational research, easy sharing of code and data, and simplifying complex setups for fellow researchers.
In large organizations and regulated industries (e.g. healthcare, finance), Docker provides essential security, agility, and compliance by isolating sensitive data and quickly deploying new services.

\textbf{Docker Compose} builds on Docker by orchestrating multiple containerized services (such as databases, backends, frontends) through simple configuration files~\cite{dockercompose2025}.
This facilitates local development of microservice architectures, comprehensive integration testing, and staging realistic production environments before deployment.

\section{Web Protocols and API Design}

\textbf{HTTP} is the foundational protocol for web and API communications, enabling clients and servers to exchange requests and responses in a standardized format, including JSON and XML payloads.

\textbf{REST APIs} are the core architectural style in web services for interoperable communication over HTTP~\cite{fielding2000rest}.
By strictly adhering to stateless client-server principles, REST APIs allow disparate systems, ML pipelines, data stores, external user interfaces, to work together seamlessly.

\section{Backend and Frontend Frameworks}

\textbf{FastAPI} is a modern web framework for building RESTful APIs with Python~\cite{fastapi2025}.
Distinctively, FastAPI leverages Python type hints and asynchronous programming, allowing developers to write concise, robust, and extremely fast endpoints.
Its native compatibility with OpenAPI and automatic documentation greatly reduces manual effort and errors.
FastAPI is widely used in both prototypes and mission-critical systems due to its speed and strong validation.

\textbf{Dash} is a Python framework for creating interactive web dashboards~\cite{dash2025}.
With minimal Python code, data scientists and engineers can build sophisticated analytics apps that integrate visualizations, user inputs, and RESTful API endpoints.
Dash leverages Flask under the hood, but developers can extend with additional Python frameworks as required for interoperable, customized solutions.

\textbf{Plotly} is an extensible visualization library available on top of Dash, useful for interactive charts, data exploration, and embedding analytics in dashboards~\cite{plotly2025}.

\section{Data Validation and Storage}

\textbf{Pydantic} is the leading Python library for data parsing and validation using type hints~\cite{pydantic2025}.
By providing strict enforcement of schemas at runtime, Pydantic allows developers to safely handle incoming data—whether from users, APIs, or databases—preventing subtle bugs and security risks.
Models inherit from BaseModel and use Python's type annotations to declaratively describe fields, their types, and validation logic.
When instantiating a model, Pydantic parses and validates input, raising structured errors for invalid data.
Its integration with FastAPI allows for automatic generation of API documentation and validation of request and response data.

\textbf{Apache Parquet} is a column-oriented data format optimized for analytical workloads in big data settings~\cite{vohra2016parquet}.
Parquet's design enables highly efficient compression and encoding, dramatically reducing storage costs and query times, especially compared to row-based formats like CSV.
